Depression and suicidality as credible signals of need in social conflicts

### 2021-06-08 Paper.rmd
Expanding section on social responses to depression

---
title: "Depression and suicidality as credible signals of need"
author:
- name: Michael R. Gaffney
  affiliation: Department of Anthropology, Washington State University
- name: Kai Adams
  affiliation: UC Berkeley Haas School of Business
- name: Kristen L. Syme
  affiliation: Department of Experimental and Applied Psychology, Vrije Universiteit Amsterdam
- name: Edward H. Hagen
  affiliation: Department of Anthropology, Washington State University
  email: edhagen@wsu.edu
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
    css: style.css
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Several theoretical models propose that depressed and suicidal individuals have inaccurate, negative perceptions of themselves and their environments [@beck1963; @joiner2007; @nolen1991]. There is strong evidence, however, that most cases of depression are caused by genuinely severe negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008; @devries2013]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. Longitudinal studies indicate that depression onsets soon after a negative event [@lewinsohn1988; @KendlerEtAl1999; @rich2005; @han2019; @sen2010] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @blekesaune2008; @metsa2013; @rosenstrom2017]. Although the relationship between negative events and depression is likely bidirectional [i.e., depression probably also causes adversity, @wichers2012], negative events predict depression even when considering only events outside of one’s control, indicating that the connection is unlikely to be driven solely by individuals who are already depressed selecting into situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events remains a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably functional instances of psychological pain and not mental dysfunctions [@Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021; @dowrick2013; @frances2013].

## Responses to depression and suicidal behavior

For depression and suicide attempts to be worth the costs involved with signal creation, meaningful social support would be required [@Hagen2003; @SymeEtAl2016]. Most studies examining responses to depression have been observational and focused on responses outside of changes in social support. In general, most report aversive responses and high levels of rejection towards the depressed, and depressed individuals are often linked to impaired social functioning [@EvraireDozois2011; @GotlibLee1989; @HirschfeldEtAl2000]. One explanation for this is that depressed individuals engage in behaviors that drive others away. In line with this reasoning, both excessive reassurance seeking and negative feedback seeking, which often accompany depression, have been shown to elicit negative emotional responses from others and increase the chance of rejection from others [@JoinerMetalsky1995; @StarrDavila2008]. Negative interactions are also seen in families as depressed children tend to report less support and more negative responses from parents than non-depressed children, although it has not been demonstrated that such interactions did not lead to depression in the first place [@MesserGross1995; @SchwartzEtAl2012].

Despite being looked for less frequently, beneficial responses to depressed individuals are also reported. Depression has been associated with reduced likelihood of aggression within families [@hops1987; @dadds1992], including those already experiencing high levels of conflict [@SheeberEtAl2001]. There is also limited evidence that depression can result in increased levels of support from others outside the family. For example, non-depressed individuals were shown to be more cooperative and exhibit more caretaking behavior towards their college depressed roommates over the course of their time living together [@HokansonEtAl1986]. Likewise, a study examining responses to confederates trained to appear ‘depressed’, ‘anxious’, or ‘normal’ found that individuals were more likely to provide advice and support to ‘depressed’ individuals than ‘anxious’ or ‘normal’ individuals despite the confederate in the ‘depression’ condition being viewed more negatively than the other conditions and the ‘depression’ condition involving similar levels of distress compared to the ‘anxious’ condition [@stephens1987].

Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for their ability to lead to beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, despite suicide rarely if ever being the main topic [@SymeEtAl2016].

@@

>>>>

---
title: "Depression and suicidality as credible signals of need in social conflicts"
author:
- name: Michael R. Gaffney
  affiliation: Department of Anthropology, Washington State University
- name: Kai Adams
  affiliation: UC Berkeley Haas School of Business
- name: Kristen L. Syme
  affiliation: Department of Experimental and Applied Psychology, Vrije Universiteit Amsterdam
- name: Edward H. Hagen
  affiliation: Department of Anthropology, Washington State University
  email: edhagen@wsu.edu
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: false
    css: style.css
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

In contrast to sadness, depression is widely believed to pathological because (1) depressed individuals putatively have inaccurate, negative perceptions of themselves and their environments [@beck1963; @joiner2007; @nolen1991], (2) social partners respond negatively, not positively, to depression [@xx], and (3) depression has many costs, increased risk of suicide paramount among them.

## Depression is probably caused by genuine adversity

Contrary to the view that the depressed have a distorted perception of their environment, there is strong evidence that most cases of depression are caused by genuinely severe negative life events, such as physical assault and death of a loved one [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008; @devries2013]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. Longitudinal studies indicate that depression onsets soon after a negative event [@lewinsohn1988; @KendlerEtAl1999; @rich2005; @han2019; @sen2010] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @blekesaune2008; @metsa2013; @rosenstrom2017]. Although the relationship between negative events and depression is likely bidirectional [i.e., depression probably also causes adversity, @wichers2012], negative events predict depression even when considering only events outside of one’s control, indicating that the connection is unlikely to be driven solely by individuals who are already depressed selecting into situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events remains a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably functional instances of psychological pain and not mental dysfunctions [@Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021; @dowrick2013; @frances2013].

## Social responses to depression

In a classic study, @Coyne1976 found that depression alienates others, a result that was subsequently confirmed in many other studies [@segrin1992; @segrin2000]. Depressed individuals self-disclose personally relevant negative issues at inappropriate times and they view such topics as more appropriate for discussion than do the non-depressed [@segrin2000]. Such self-disclosures have been shown to be a key ingredient in the rejection of depressed persons by others, and "may appropriately be understood as an attempt to elicit social support from targets" [@segrin1994, p. 657]. In interactions with spouses, depressed individuals express anger and aggression, and make frequent demands for help [@segrin1994]. Excessive reassurance seeking -- repeatedly requesting reassurance that one is lovable and worthy despite previous attempts by others to provide such reassurance -- is another factor implicated in the rejection of the depressed by others [@joiner1999; @JoinerMetalsky1995; @StarrDavila2008]. Beneficial responses to depressed individuals have also been reported, such as increased caretaking [@HokansonEtAl1986], more offers of advice and support [@stephens1987], and reduced aggression within families [@hops1987; @dadds1992; @SheeberEtAl2001].

Negative social responses to depression are widely interpreted as caused by impaired social functioning in the depressed [@EvraireDozois2011; @GotlibLee1989; @HirschfeldEtAl2000], including by evolutionary theorists of depression [@nettle2004; @allen2006; @nesse2006; @myers2016]. These theorists have overlooked the role of conflict in depression.

<!-- including those already experiencing high levels of conflict []. There is also limited evidence that depression can result in increased levels of support from others outside the family. For example, non-depressed individuals were shown to be more cooperative and exhibit more caretaking behavior towards their college depressed roommates over the course of their time living together [@HokansonEtAl1986]. Likewise, a study examining responses to confederates trained to appear ‘depressed’, ‘anxious’, or ‘normal’ found that individuals were more likely to provide advice and support to ‘depressed’ individuals than ‘anxious’ or ‘normal’ individuals despite the confederate in the ‘depression’ condition being viewed more negatively than the other conditions and the ‘depression’ condition involving similar levels of distress compared to the ‘anxious’ condition . -->

<!-- One explanation for this is that depressed individuals engage in behaviors that drive others away. In line with this reasoning, both excessive reassurance seeking and negative feedback seeking, which often accompany depression, have been shown to elicit negative emotional responses from others and increase the chance of rejection from others [@JoinerMetalsky1995; @StarrDavila2008]. Negative interactions are also seen in families as depressed children tend to report less support and more negative responses from parents than non-depressed children, although it has not been demonstrated that such interactions did not lead to depression in the first place [@MesserGross1995; @SchwartzEtAl2012]. -->

<!-- After confirming Coyne's (1976a, 1976b) finding that depression alienates others, including strangers in brief interactions (for a meta-analysis, see Segrin, 1992) researchers began to explore what it is that depressed people do or fail to do that leads to others' hostility and rejection. A large literature has since accumulated documenting differences in interpersonal interactions by depressed and non-depressed individuals. Findings include differences in communication ‘style,' such as vocal characteristics, eye gaze, and gestures, and differences in ‘content,' such as preferred topics and themes of conversation (Segrin, 1994; Segrin, 2000). -->

<!-- With regard to style, e.g., paralinguistic qualities like low rate, low volume, and low monotone pitch, depressed people appear to be ''very 'skilled' at communicating sadness and despair'' (Segrin, 2000, p. 385). With regard to content, depressed individuals are eager to self-disclose personally relevant negative issues, often at what are judged to be inappropriate times, and they view such topics as more appropriate for discussion than do the non-depressed (Segrin, 2000). Such self-disclosures have been shown to be a key ingredient in the rejection of depressed persons by others, and ''may appropriately be understood as an attempt to elicit social support from targets'' (Segrin, 1994, p. 657). In interactions with spouses, depressed individuals express anger and aggression, and make frequent demands for help. Segrin (1994) speculates that  -->

<!-- The discussion of negative well-being, dysphoric feelings, and frequent demands for help may impose on other people, making them feel either responsible or obligated to offer assistance or find out the cause of the depression. This feeling of obligation or responsibility could potentially prove to be aversive and motivate withdrawal and rejection of the depressed person. (p. 660) -->

<!-- Excessive reassurance seeking (ERS) is another well-studied factor implicated in depression and rejection by others. ERS is defined as repeatedly requesting reassurance that one is lovable and worthy despite previous attempts by others to provide such reassurance. According to Joiner (1999) -->

<!-- The key idea is that the aversive properties of depression are not interpersonally aversive unless clearly and repeatedly signaled to other people and, further, signaled in such a way that others are both implicated in the development of the problem (e.g., ''you don't really love me anymore, do you?"), as well as overwhelmed by demands, whether implicit or stated, to solve the problem. (p. 272) -->

<!-- Our view is that excessive reassurance-seeking predisposes people to the development of depressive symptoms. Initially, this may occur as a function of growing demoralization of the high reassurance-seeking person, who has struggled to gain -- but has been disappointed with -- others' responsiveness. (p. 273) -->

<!-- A meta-analysis of 38 studies found ERS to be moderately associated with depression (r = .38) and weakly but significantly associated with rejection (r = .14), although the relationship with rejection could be stronger when the seeker is depressed (Starr, 2008). -->

<!-- Because depression alienates others, however, differences in the style and content of interpersonal interactions of the depressed vs. non-depressed, rather than being seen as an aversive signal of need, are interpreted as deficits, such as social skill deficits or 'excessive' reassurance-seeking (Segrin, 1994; Joiner et al., 1999). Yet infant cries, clear signals of need, motivate caregiving, in part by being so aversive that caregivers work to terminate the cries. The noxiousness of these cries can also evoke abuse or avoidance by caregivers (Owings & Zeifman, 2004).  -->

<!-- Normal sadness can also alienate others. Young children, for instance, find sad people to be rather unlikeable (Glasberg, 1981; Glasberg, 1982), and report controlling their own sadness, especially around peers, because they fear negative interpersonal consequences (Zeman, 1996). Adults commonly regulate their emotions too, including sadness, in part because there are negative social consequences to emotional expression, in addition to positive ones (for reviews see Gross, 1998; Gross, 2002; Ochsner, 2005).  -->

@@

<!-- Need to integrate this -->
Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for their ability to lead to beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, despite suicide rarely if ever being the main topic [@SymeEtAl2016].
